---
title: "Phylodynamics Practical"
author: "Simon Frost"
date: "1 December 2015"
output: html_document
runtime: shiny
---

```{r, echo=FALSE}
# Load  libraries
suppressPackageStartupMessages(library(shiny))
suppressPackageStartupMessages(library(ape))
suppressPackageStartupMessages(library(phangorn))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(adephylo))
suppressPackageStartupMessages(library(magrittr))
suppressWarnings(suppressMessages(library(ggtree)))
suppressWarnings(suppressMessages(library(gridExtra)))
```

```{r, echo=FALSE}
RLchronos <- function (phy, lambda = 1, model = "correlated", quiet = FALSE, 
                       calibration = makeChronosCalib(phy),
                       control = chronos.control()) 
{
  model <- match.arg(tolower(model), c("correlated", "relaxed", 
                                       "discrete"))
  n <- Ntip(phy)
  ROOT <- n + 1L
  m <- phy$Nnode
  el <- phy$edge.length
  if (any(el < 0)) 
    stop("some branch lengths are negative")
  e1 <- phy$edge[, 1L]
  e2 <- phy$edge[, 2L]
  N <- length(e1)
  TIPS <- 1:n
  EDGES <- 1:N
  tol <- control$tol
  node <- calibration$node
  age.min <- calibration$age.min
  age.max <- calibration$age.max
  if (model == "correlated") {
    basal <- which(e1 == ROOT)
    Nbasal <- length(basal)
    ind1 <- EDGES[-basal]
    ind2 <- match(e1[EDGES[-basal]], e2)
  }
  age <- numeric(n + m)
  if (!quiet) 
    cat("\nSetting initial dates...\n")
  
  ## seq.nod is sorted by node index.
  seq.nod <- .Call(seq_root2tip, phy$edge, n, phy$Nnode)
  ii <- 1L
  repeat {
    ini.time <- age
    ini.time[ROOT:(n + m)] <- NA
    ini.time[node] <- if (is.null(age.max)) 
      age.min
    else runif(length(node), age.min, age.max)
    if (is.na(ini.time[ROOT])) 
      ini.time[ROOT] <- if (is.null(age.max)) 
        3 * max(age.min)
      else 3 * max(age.max)
    ISnotNA.ALL <- unlist(lapply(seq.nod, function(x) sum(!is.na(ini.time[x]))))
    
    ## o is a vector of tip indices, sorted by those which have the
    ## most non-NA nodes between them and the root.
    o <- order(ISnotNA.ALL, decreasing = TRUE)
    ## seq.nod[o] is the vector of nodes visited on the path from
    ## root to tip, starting with the root and ending with the
    ## tip.  We need to modify it so that it handles the more
    ## recent tips first.
    
    not.NA.tip.ages <- ini.time[o]
    
    tips.by.age <- order(not.NA.tip.ages, decreasing=TRUE)
    
    for (y in seq.nod[tips.by.age]) {
      ## Are any un-set still?
      ISNA <- is.na(ini.time[y])
      if (any(ISNA)) {
        ## i == 1 is the root, and we already set that.
        i <- 2L
        while (i <= length(y)) {
          if (ISNA[i]) {
            ## if the ith node on this path is unset, then find the
            ## next *set* node; j is its index.
            j <- i + 1L
            while (ISNA[j]) j <- j + 1L
            ## nb.val is now the number of unset nodes on this path.
            nb.val <- j - i
            ## by is the interval of time between the date-set nodes
            ## divided by the number of edges on the path between
            ## them (which is the number of nodes + 1).
            by <- (ini.time[y[i - 1L]] - ini.time[y[j]])/(nb.val + 1)
            ini.time[y[i:(j - 1L)]] <-
              ini.time[y[i - 1L]] - by * seq_len(nb.val)
            i <- j + 1L
          }
          else i <- i + 1L
        }
      }
    }
    if (all(ini.time[e1] - ini.time[e2] >= 0)) 
      break
    ii <- ii + 1L
    if (ii > 1000) 
      stop("cannot find reasonable starting dates after 1000 tries:\nmaybe you need to adjust the calibration dates")
  }
  if(!quiet){
  cat("Initial dates set.\n")
  }
  ini.rate <- el/(ini.time[e1] - ini.time[e2])
  if (model == "discrete") {
    Nb.rates <- control$nb.rate.cat
    minmax <- range(ini.rate)
    if (Nb.rates == 1) {
      ini.rate <- sum(minmax)/2
    }
    else {
      inc <- diff(minmax)/Nb.rates
      ini.rate <- seq(minmax[1] + inc/2, minmax[2] - inc/2, 
                      inc)
      ini.freq <- rep(1/Nb.rates, Nb.rates - 1)
      lower.freq <- rep(0, Nb.rates - 1)
      upper.freq <- rep(1, Nb.rates - 1)
    }
  } else Nb.rates <- N
  
  
  ## The internal nodes are numbered (n+1), ..., (n+m).
  unknown.ages <- 1:m + n
  ## These appear to only be for internal nodes.
  lower.age <- rep(tol, m)
  upper.age <- rep(1/tol, m)
  
  bdd.internal.node.indices <- which(node > n)
  bdd.internal.nodes <- node[bdd.internal.node.indices]
  lower.age[bdd.internal.nodes - n] <- age.min[bdd.internal.node.indices]
  upper.age[bdd.internal.nodes - n] <- age.max[bdd.internal.node.indices]
  
  ## Eliminate nodes where age.min != age.max; whatever's left,
  ## we can set the age from age.min.
  ii <- which(age.min != age.max)
  ## if (length(ii)) {
  ##     node <- node[-ii]
  ##     age.min <- age.min[-ii]
  ## }
  if (length(node[-ii])) {
    age[node[-ii]] <- age.min[-ii]
  } else {
    age[node] <- age.min
  }
  
  
  ## Get rid of the internal nodes that we have knowledge of.
  fixed.internal.nodes <-
    bdd.internal.nodes[age.min[bdd.internal.node.indices] ==
                       age.max[bdd.internal.node.indices]]
  if (length(fixed.internal.nodes)) {
    unknown.ages <- unknown.ages[n - fixed.internal.nodes]
    lower.age <- lower.age[n - fixed.internal.nodes]
    upper.age <- upper.age[n - fixed.internal.nodes]
  }
  
  known.ages <- c(TIPS, bdd.internal.nodes)
  lower.rate <- rep(tol, Nb.rates)
  upper.rate <- rep(100 - tol, Nb.rates)
  degree_node <- tabulate(phy$edge)
  eta_i <- degree_node[e1]
  eta_i[e2 <= n] <- 1L
  X <- vector("list", N)
  for (i in EDGES) {
    j <- integer()
    if (e1[i] != ROOT) 
      j <- c(j, which(e2 == e1[i]))
    if (e2[i] >= n) 
      j <- c(j, which(e1 == e2[i]))
    X[[i]] <- j
  }
  ## List of indices in e2 of nodes with unknown ages.
  D_ki <- match(unknown.ages, e2)
  ## List of indices in e1 of edges coming out of each of the
  ## unknown nodes.
  A_ki <- lapply(unknown.ages, function(x) which(x == e1))
  gradient.poisson <- function(rate, node.time) {
    age[unknown.ages] <- node.time
    real.edge.length <- age[e1] - age[e2]
    ## el is the list of edge lengths in the tree.
    gr <- el/rate - real.edge.length
    tmp <- el/real.edge.length - rate
    gr.dates <- sapply(A_ki, function(x) sum(tmp[x])) - tmp[D_ki]
    c(gr, gr.dates)
  }
  gradient <- switch(model, correlated = function(rate, node.time) {
    gr <- gradient.poisson(rate, node.time)
    gr[RATE] <- gr[RATE] - lambda * 2 * (eta_i * rate -
                                         sapply(X, 
                                                function(x) sum(rate[x])))
    if (Nbasal == 2) {
      i <- basal[1]
      j <- basal[2]
      gr[i] <- gr[i] - lambda * (rate[i] - rate[j])
      gr[j] <- gr[j] - lambda * (rate[j] - rate[i])
    } else {
      for (i in 1:Nbasal) j <- basal[i]
      gr[j] <- gr[j] - lambda * 2 * (rate[j] * (1 - 1/Nbasal) - 
                                     sum(rate[basal[-i]])/Nbasal)/(Nbasal - 1)
    }
    gr
  }, relaxed = function(rate, node.time) {
    gr <- gradient.poisson(rate, node.time)
    mean.rate <- mean(rate)
    gr[RATE] <- gr[RATE] + lambda * 2 * dgamma(rate, mean.rate) * 
      (rank(rate)/Nb.rates - pgamma(rate, mean.rate))
    gr
  }, discrete = NULL)
  log.lik.poisson <- function(rate, node.time) {
    age[unknown.ages] <- node.time
    real.edge.length <- age[e1] - age[e2]
    if (isTRUE(any(real.edge.length < 0))) 
      return(-1e+100)
    B <- rate * real.edge.length
    sum(el * log(B) - B - lfactorial(el))
  }
  penal.loglik <-
    switch(model,
           correlated = function(rate, node.time)
           {
             loglik <- log.lik.poisson(rate, node.time)
             if (!is.finite(loglik)) return(-1e+100)
             loglik - lambda * (sum((rate[ind1] - rate[ind2])^2) + 
                                var(rate[basal]))
           },
           relaxed = function(rate, node.time)
           {
             loglik <- log.lik.poisson(rate, node.time)
             if (!is.finite(loglik)) return(-1e+100)
             mu <- mean(rate)
             loglik - lambda * sum((1:N/N - pgamma(sort(rate), mean(rate)))^2)
           },
           discrete = if (Nb.rates == 1) function(rate, node.time) log.lik.poisson(rate, 
                            node.time) else function(rate, node.time, freq) {
                              if (isTRUE(sum(freq) > 1)) return(-1e+100)
                              rate.freq <- sum(c(freq, 1 - sum(freq)) * rate)
                              log.lik.poisson(rate.freq, node.time)
                            })
  opt.ctrl <- list(eval.max = control$eval.max, iter.max = control$iter.max)

  ## In the optimization, the rates come first and the ages come second.
  ## p is the vector of parameters.
  ## RATE and AGE are the corresponding indices in the parameter;
  ## LOW and UP are bounds on the parameters.
  RATE <- 1:Nb.rates
  AGE <- Nb.rates + 1:length(unknown.ages)
  if (model == "discrete") {
    if (Nb.rates == 1) {
      start.para <- c(ini.rate, ini.time[unknown.ages])
      f <- function(p) -penal.loglik(p[RATE], p[AGE])
      g <- NULL
      LOW <- c(lower.rate, lower.age)
      UP <- c(upper.rate, upper.age)
    }
    else {
      FREQ <- length(RATE) + length(AGE) + 1:(Nb.rates - 1)
      start.para <- c(ini.rate, ini.time[unknown.ages], 
                      ini.freq)
      f <- function(p) -penal.loglik(p[RATE], p[AGE], p[FREQ])
      g <- NULL
      LOW <- c(lower.rate, lower.age, lower.freq)
      UP <- c(upper.rate, upper.age, upper.freq)
    }
  } else {
    start.para <- c(ini.rate, ini.time[unknown.ages])
    f <- function(p) -penal.loglik(p[RATE], p[AGE])
    g <- function(p) -gradient(p[RATE], p[AGE])
    LOW <- c(lower.rate, lower.age)
    UP <- c(upper.rate, upper.age)
  }
  k <- length(LOW)
  if (!quiet) 
    cat("Fitting in progress... get a first set of estimates\n")
  out <- nlminb(start.para, f, g, control = opt.ctrl, lower = LOW, 
                upper = UP)
  if (model == "discrete") {
    if (Nb.rates == 1) {
      f.rates <- function(p) -penal.loglik(p, current.ages)
      f.ages <- function(p) -penal.loglik(current.rates, p)
    }
    else {
      f.rates <- function(p) -penal.loglik(p, current.ages, 
                                           current.freqs)
      f.ages <- function(p) -penal.loglik(current.rates, 
                                          p, current.freqs)
      f.freqs <- function(p) -penal.loglik(current.rates, 
                                           current.ages, p)
      g.freqs <- NULL
    }
    g.rates <- NULL
    g.ages <- NULL
  } else {
    f.rates <- function(p) -penal.loglik(p, current.ages)
    g.rates <- function(p) -gradient(p, current.ages)[RATE]
    f.ages <- function(p) -penal.loglik(current.rates, p)
    g.ages <- function(p) -gradient(current.rates, p)[AGE]
  }
  current.ploglik <- -out$objective
  current.rates <- out$par[RATE]
  current.ages <- out$par[AGE]
  if (model == "discrete" && Nb.rates > 1) 
    current.freqs <- out$par[FREQ]
  dual.iter.max <- control$dual.iter.max
  i <- 0L
  if (!quiet) 
    cat("         Penalised log-lik =", current.ploglik, 
        "\n")
  repeat {
    if (dual.iter.max < 1) 
      break
    if (!quiet) 
      cat("Optimising rates...")
    out.rates <- nlminb(current.rates, f.rates, g.rates, 
                        control = list(eval.max = 1000, iter.max = 1000, 
                          step.min = 1e-08, step.max = 0.1), lower = lower.rate, 
                        upper = upper.rate)
    new.rates <- out.rates$par
    if (-out.rates$objective > current.ploglik) 
      current.rates <- new.rates
    if (model == "discrete" && Nb.rates > 1) {
      if (!quiet) 
        cat(" frequencies...")
      out.freqs <- nlminb(current.freqs, f.freqs,
                          control = list(eval.max = 1000, 
                            iter.max = 1000, step.min = 0.001, step.max = 0.5), 
                          lower = lower.freq, upper = upper.freq)
      new.freqs <- out.freqs$par
    }
    if (!quiet) 
      cat(" dates...")
    out.ages <- nlminb(current.ages, f.ages, g.ages,
                       control = list(eval.max = 1000, 
                         iter.max = 1000, step.min = 0.001, step.max = 100), 
                       lower = lower.age, upper = upper.age)
    new.ploglik <- -out.ages$objective
    if (!quiet) 
      cat("", current.ploglik, "\n")
    if (new.ploglik - current.ploglik > 1e-06 && i <= dual.iter.max) {
      current.ploglik <- new.ploglik
      current.rates <- new.rates
      current.ages <- out.ages$par
      if (model == "discrete" && Nb.rates > 1) 
        current.freqs <- new.freqs
      out <- out.ages
      i <- i + 1L
    }
    else break
  }
  if (!quiet) 
    cat("\nDone.\n")
  if (model == "discrete") {
    rate.freq <- if (Nb.rates == 1) 
      current.rates
    else mean(c(current.freqs, 1 - sum(current.freqs)) * 
              current.rates)
    logLik <- log.lik.poisson(rate.freq, current.ages)
    PHIIC <- list(logLik = logLik, k = k, PHIIC = -2 * logLik + 
                  2 * k)
  } else {
    logLik <- log.lik.poisson(current.rates, current.ages)
    PHI <- switch(model,
                  correlated = (current.rates[ind1] - 
                                current.rates[ind2])^2 + var(current.rates[basal]), 
                  relaxed = (1:N/N - pgamma(sort(current.rates), mean(current.rates)))^2)
    PHIIC <- list(logLik = logLik, k = k, lambda = lambda, 
                  PHIIC = -2 * logLik + 2 * k + lambda * svd(PHI)$d)
  }
  ## DEBUGGING
  ## attr(phy, "call") <- "FOOBAR"
  attr(phy, "call") <- match.call()
  attr(phy, "ploglik") <- -out$objective
  attr(phy, "rates") <- current.rates
  if (model == "discrete" && Nb.rates > 1) 
    attr(phy, "frequencies") <- current.freqs
  attr(phy, "message") <- out$message
  attr(phy, "PHIIC") <- PHIIC
  age[unknown.ages] <- current.ages
  phy$edge.length <- age[e1] - age[e2]
  class(phy) <- c("chronos", class(phy))
  phy
}
```


```{r,echo=FALSE}
busyIndicator <- function(text = "Calculation in progress..",img = "shinysky/busyIndicator/ajaxloaderq.gif", wait=1000) {
	tagList(
  		singleton(tags$head(
    		tags$link(rel="stylesheet", type="text/css",href="shinysky/busyIndicator/busyIndicator.css")
  			))
  		,div(class="shinysky-busy-indicator",p(text),img(src=img))
  		,tags$script(sprintf(
  		"	setInterval(function(){
  		 	 if ($('html').hasClass('shiny-busy')) {
  		    setTimeout(function() {
  		      if ($('html').hasClass('shiny-busy')) {
  		        $('div.shinysky-busy-indicator').show()
  		      }
  		    }, %d)  		    
  		  } else {
  		    $('div.shinysky-busy-indicator').hide()
  		  }
  		},100)
  		",wait)
  			)
	)	
}
```


## Loading the tree

First of all, we have to load a tree. Use the dialogue below to choose a file that contains a tree in Newick format.

```{r, echo=FALSE}
inputPanel(
  fluidPage(
    fileInput("nwk","Choose tree file")
  )
)
renderPrint({
  tr()
})
renderPlot({
  if(length(input$nwk$datapath)==0) return(invisible())
  g <- ggtree(tr(),layout="unrooted")
  g <- g + ggtitle("Unrooted")
  g
})
renderPlot({
  if(length(input$nwk$datapath)==0) return(invisible())
  tree <- tr()
  mrd <- max(distRoot(tree))
  g <- ggtree(tree,showDistance=TRUE) + scale_x_continuous(limits=c(0,mrd*1.5))
  g <- g +ggtitle("Rooted")
  g + geom_text(aes(label=label),size=2,hjust=-0.05)
})
```

```{r, echo=FALSE}
tr <- reactive({
    if(length(input$nwk$datapath)==0) return(invisible())
    read.tree(input$nwk$datapath[1])
})
```


## Loading annotations

Secondly, you need to load in annotations; for this practical, annotations comprise of the sampling time and a grouping variable. Use the dialogue below to upload the annotations file.

```{r, echo=FALSE}
inputPanel(
  fluidPage(
    fileInput("annotation","Choose annotation file")
  )
)
```

Here is a summary of the annotations.

```{r,echo=FALSE}
fluidPage(
  renderPrint({
    if(length(input$annotation$datapath)==0) return(invisible())
    summary(annotations())
  })
)
```

Here is a more detailed view of the annotations.

```{r, echo=FALSE}
fluidPage(
  renderDataTable({
    annotations()
  })
)
```

```{r, echo=FALSE}
annotations <- reactive({
    if(length(input$annotation$datapath)==0) return()
    tbl <- read.table(input$annotation$datapath[1],header=T,row.names=NULL,sep="\t")
})
```


## Distances from a tree

We can calculate the distances (in terms of genetic divergence) between sequences from the phylogenetic tree. This, along with the tree itself, can sometimes help to indicate levels of clustering in the data.

```{r,echo=FALSE}
actionButton("calcdist","Calculate distances")
```

A histogram of the distances.

```{r,echo=FALSE}
renderPlot({
  if(input$calcdist==0) return()
  d <- dst()
  d <- d[lower.tri(d)]
  hist(d)
})
```

A summary of the distances.

```{r, echo=FALSE}
renderPrint({
  if(input$calcdist==0) return(invisible())
  d <- dst()
  d <- d[lower.tri(d)]
  summary(d)
})
```

```{r,echo=FALSE}
dst <- reactive({
  if(input$calcdist==0) return()
  cophenetic.phylo(tr())
})
```


## Root-to-tip regression

The phylogeny, as loaded into the program, may be rooted arbitrarily. To root the tree on the basis of sampling time, we can perform a *root-to-tip* regression, in which we try to find the root that gives the best straight line fit between root-to-tip distances and sampling time. This gives us an estimate of:

- The time to the most recent common ancestor (TMRCA)
- The rate of evolution

```{r,echo=FALSE}
actionButton("rtt","Root-to-tip regression")
treertt <- reactive({
  if(input$rtt==0) return()
  annot <- annotations()
  tr.rtt <- tr()
  tr.rtt <- multi2di(tr.rtt)
  tr.rtt$edge.length <- tr.rtt$edge.length+1e-6
  td <- as.double(annot[match(tr.rtt$tip.label,annot[,1]),2])
  tr.rtt <- rtt(tr.rtt,td,objective="rsquared")
  tr.rtt
})
treelm <- reactive({
  if(input$rtt==0) return()
  tr.rtt <- treertt()
  rd <- distRoot(tr.rtt)
  annot <- annotations()
  td <- as.double(annot[match(tr.rtt$tip.label,annot[,1]),2])
  rtt.lm <- lm(rd~td)
  root.time <- unname(-as.double(coef(rtt.lm)[1])/coef(rtt.lm)[2])
  list(root.time,rd,td,rtt.lm)
})
```

This plot illustrates the observed and predicted root-to-tip distances based on sampling time.

```{r, echo=FALSE}
renderPlot({
  if(input$rtt==0) return()
  tr.lm <- treelm()
  rd <- tr.lm[[2]]
  td <- tr.lm[[3]]
  root.time <- tr.lm[[1]]
  plot(rd~td,xlab="Time",ylab="Root to tip distance",ylim=c(0,max(rd)),xlim=c(root.time,max(td)),pch=16,col="red")
  abline(tr.lm[[4]])
  abline(h=0,lty=2)
})
```

Here is the summary of the root-to-tip regression.

```{r,echo=FALSE}
renderPrint({
  if(input$rtt==0) return(invisible())
  summary(treelm()[[4]])
})
```

The TMRCA can be calculated from this regression.

```{r, echo=FALSE}
renderPrint({
  if(input$rtt==0) return(invisible())
  tlm <- treelm()
  paste("The TMRCA is",tlm[[1]])
})
```


## 'Clockifying' a tree

The next step is to make the tree actually follow a clock. The button below will fit a strict clock, then plot the groups onto the tree.

```{r,echo=FALSE}
inputPanel(
  numericInput("roottime","Root time", 0),
  actionButton("clockify","Clockify!")
)
renderPlot({
  if(input$clockify==0) return()
  tree.dated <- clockify()
  dated.tree <- read.tree(text=write.tree(tree.dated))
  annot <- annotations()
  g <- ggtree(dated.tree)+theme_tree2()
  g <- g %<+% annot
  g + geom_tippoint(aes(color=Group),size=2)+theme(legend.position="right")
})
```

```{r, echo=FALSE}
clockify <- reactive({
  if(input$clockify==0) return()
    tr.rtt <- treertt()
    root.time <- as.double(input$roottime)
    calibrating.values <- makeChronosCalib(tr.rtt)
    annot <- annotations()
    td <- as.double(annot[match(tr.rtt$tip.label,annot[,1]),2])
    max.time <- max(td)
    calibrating.values$age.min <- max.time - root.time
    calibrating.values$age.max <- max.time - root.time
    calibrating.values <- rbind(calibrating.values,
                            data.frame(node=seq(1,length(td)),
                                       age.min=max.time - td,
                                       age.max=max.time - td,
                                       soft.bounds=FALSE))

    dated.tree <- RLchronos(tr.rtt, 
                     lambda=1, 
                     model="discrete", 
                     calibration=calibrating.values,
                     control=chronos.control(nb.rate.cat=1))
    dated.tree
})
```
